\subsection{Minimal variational requirements}

We seek a variational formulation for the alignment field $\phi(\theta)$ satisfying the following conditions:
\begin{itemize}
\item invariance under smooth reparametrizations of $\ThetaMan$,
\item locality on the statistical manifold,
\item minimal derivative order,
\item absence of any geometric structure beyond the Fisher--Rao metric.
\end{itemize}

The Fisher--Rao metric uniquely determines the invariant volume element
$\sqrt{\det G}\,\mathrm{d}^D\theta$ and the Laplace--Beltrami operator
\begin{equation}
\Delta_G
=
\frac{1}{\sqrt{\det G}}\,
\partial_i\!\left(
\sqrt{\det G}\,G^{ij}\partial_j
\right).
\end{equation}

No additional geometric tensors, background structures, or intrinsic scales are introduced.

We emphasize that no explicit curvature couplings (such as $\phi\,R(G)$) are included.
While such terms are compatible with reparametrization invariance, they introduce
additional geometric structure beyond that strictly required to encode empirical
alignment.
The present construction is intentionally restricted to the minimal scalar sector
generated solely by the Fisher--Rao metric and its associated Laplace--Beltrami operator.

\subsection{Empirically coupled action}

The minimal action coupling the alignment field to empirical data is
\begin{equation}
\boxed{
S[\phi;q]
=
\int_{\ThetaMan} \mathrm{d}^D\theta\,\sqrt{\det G}
\left[
\frac{1}{2}\,G^{ij}\partial_i\phi\,\partial_j\phi
+
\gamma\,A(\theta;q)\,\phi
\right].
}
\label{eq:ActionPoisson}
\end{equation}

This action is uniquely fixed by the requirements of locality on $\ThetaMan$,
reparametrization invariance, minimal derivative order, and the absence of
additional geometric structure beyond the Fisher--Rao metric. In particular,
the linear coupling $A(\theta;q)\,\phi$ is the only admissible lowest-order
interaction between the empirical source and the scalar field. Higher-order
or nonlinear couplings would introduce non-minimal empirical scales without
improving interpretability or geometric consistency.

The coupling constant $\gamma$ fixes only the overall normalization of the
auxiliary field $\phi$ and introduces no intrinsic scale.
Since $\phi$ is not an observable quantity, $\gamma$ carries no independent
geometric or statistical content and may be absorbed into a rescaling of $\phi$.
Its explicit presence is retained solely for notational clarity when comparing
different empirical sources.

This action defines a purely geometric relaxation theory on the statistical
manifold. The alignment field $\phi$ is an auxiliary variable whose equilibrium
configuration encodes a smooth, Fisher--geometrically consistent response to
empirical deformation.
